Does personal relevance attenuate the effect of government branding? Evidence from a survey experiment on air-pollution policy

### 2019-07-18 paper3_relevance and symbols_revision_01.Rmd
Add files via upload

Public Administration scholars have recently shown a growing interest in government branding, and its potency to influence citizens' perceptions and steer their behaviour[@eshuis_2012; @marland_2016; @marland_2017; @zavattaro_website_2018; @sataen_2015; @karens_2016; @alonbarkat_2017; @teodoro_2018]. Recent experimental studies established that symbolic, brand elements[^footnote_branding] entangled in government communications (e.g. logos, brand names, figures, celebs) can have an emotional effect on citizens, which may significantly enhance their trust in government organizations and policies [@karens_2016; @alonbarkat_2017; @teodoro_2018], and positively affect their evaluation of services [@marvel_2015a]. This scholarly interest corresponds with an overall increase in the use of branding by public-sector organizations [@eshuis_2012; ].

@@

The employment of symbolic communications by government organizations can be perceived favorably, insofar as it enables organizations to engage more effectively with citizens and overcome their negative prior beliefs and biases against the public sector [@marvel_2015a; @marvel_2015; @hvidman_2016]. Accordingly, it can mitigate citizens' distrust in government organizations and enhance their collaboration with their policies. On the other hand, one can argue that the persuasion of citizens through symbols is problematic, since the emotive effect of symbols undermines the fundamental normative expectation that citizens' evaluation of their government should be based on their critical judgment. Furthermore, symbols can compensate for organizations' poor performance and for logically unpersuasive informative content, as empirically demonstrated by previous studies [@alonbarkat_2017; @alonbarkat_2018]. They can yield a misalignment between citizens' positive perceptions and the "real world". This raises the concern that government organizations will strategically exploit symbols to manipulate public opinion and mitigate justified public criticism of their failures.

The latter concern is the main motivation of this study. Accordingly, in this study I explore the boundaries of governments' persuasion (and possibly manipulation) of citizens through symbolic communication, to asses the scope of the risk that symbolic communication practices pose for democratic accountability. Specifically, I explore whether citizens are susceptible to symbolic communication also when the issue at hand is highly relevant for them, and may have a salient impact on their lives. This is particularly important, since we tend to expect these "vulnerable" groups, both positively and normatively, to be aware of government actions that concern their interests, and hold them to account. That positive expectation can be supported by the social psychology theory. The Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011], postulates that people are more likely to focus on the substantive content of the communication and ignore "peripheral" symbolic elements, the more they perceive the policy issue at hand as having personal relevance for themselves.

Employing a survey experiment, I test this theoretical expectation in the context of the Israeli Environment Protection Ministry, and its policy plan for reducing the air pollution in a specific geographical area (the Haifa Bay). The population of the study (*n* = `r SVIVA2_01 %>% nrow()`) consists of citizens residing in the polluted area, matched with residents of other cities in the center of the country that are not exposed to air-pollution. Participants are presented with a policy plan regarding the air-pollution in that area, while experimentally varying both its appearance (familiar symbolic elements from the Ministry's communications versus two control groups) and content (logically versus logical plan). I examine the effects of these manipulations on participants' trust in the policy plan in interaction with the two areas, to test whether residents of the polluted area are less affected by familiar symbolic elements (and more affected by the different in the substantive content). I also test the moderating role of personal relevance experimentally by priming participants to think about accut environmental policy issues in their area of residence. Contrary to my expectations, The results indicate that even high levels of perceived personal relevance do not significantly attenuate the effect of symbols on citizens' trust. These findings tend to suggest that the boundaries of the effect of symbolic communication are wider than expected. I discuss the theoretical and normative implications of these findings.

@@

The Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo, provides a comprehensive theoretical framework for investigating the effects of communications on people's attitudes [@petty_1986; @petty_2011]. According to ELM, people may respond differently to a communication, depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on a message, they are likely to focus on the arguments in the communication that are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route"). These two routes for persuasion are regarded as two ends of an "elaboration continuum". People may also be located between these two ends of the continuum, and then a mixture of these processes will determine their persuasion.

>>>>

Public Administration scholarship has recently shown a growing interest in public sector branding, and its potency to influence citizens' perceptions and direct their behaviour [@eshuis_2012; @marland_2016; @marland_2017; @zavattaro_website_2018; @sataen_2015]. Behavioural studies established that symbolic, brand elements[^footnote_branding] entangled in government communications (e.g. brand names, logos, colors and figures) can have an emotional effect on citizens, which leads them to view public organizations and their policies and services more favourably [@marvel_2015a; @karens_2016; @alonbarkat_2017; @teodoro_2018]. This growing scholarly interest corresponds with the overall increase in the use of branding practices by public-sector organizations over the past decade, in part due to the expension of social media and ditigal communication platforms [@eshuis_2012; @marland_2017; @mickoleit_2014; @mergel_2013].

@@

The ability of government organizations to affect citizens via symbolic communications can be perceived favorably, insofar as it enables organizations to engage more effectively with citizens and overcome their unjustified negative prior beliefs and biases against the public sector [@marvel_2015a; @marvel_2015; @hvidman_2016; @hvidman_2019]. Accordingly, it can mitigate citizens' distrust in government organizations and underevaluation of their performance, which represent a serious concern.

Nonetheless, the persuasion of citizens through symbols is also highly problematic. Normatively, we expect citizens in a democratic society to form their opinions about their government based on their critical thinking, rather then on uncontious emotive responses, produced by government agencies themselves. A yet more serious concern regards the potency of symbolic elements in government communications to compensate for organizations' poor performance, poorly planned policy plans and logically unpersuasive explanations, as indicated by previous studies [@alonbarkat_2017; @alonbarkat_2018]. They can yield a misalignment between citizens' positive perceptions and the "real world", which would undermine the democratic accountability and responsiveness of public organizations. Moreover, government organizations (or politicians controlling them) could strategically exploit such symbolic communication to manipulate public opinion and mitigate justified public criticism of agencies' failures.

The abovementioned concerns of the negative implications of branding are the main motivation of this study. These concerns necessitate a better understanding of the boundaries of governments' persuasion (and possibly manipulation) of citizens through symbolic communication. Understanding the limitations of symbolic persuasion would enable to asses the scope of the risk that these practices pose for democracy.

For this purpose, in this study I explore whether the effect of symbols is restricted to policy issues that are not perceived by citizens as having a significant consequence for their personal lives, or rather, that citizens can be similarly persuaded by symbols even with regard to such personally relevant issues. This question is particularly important, since we tend to expect these affected groups to pay more attention to government actions that concern their interests, and to hold them to account for their consequences to a greater extent. That expectation can be supported, theoretically, by the social psychology theory Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011]. The model postulates that people are more likely to focus on the substantive content of the communication and ignore "peripheral" symbolic elements, the more they perceive the policy issue at hand as having personal relevance for themselves.

Employing a survey experiment, I test this latter theoretical expectation in the context of the Israeli Environment Protection Ministry, and its policy plan for reducing the air pollution in a specific geographical area -- the Haifa Bay. The research sample (*n* = `r SVIVA2_01 %>% nrow()`) consists of citizens residing in the polluted area, matched with residents of other cities in the center of the country that are not exposed to air-pollution. I present all participants a policy plan regarding the air-pollution in the Haifa Bay area, while experimentally varying both its appearance (familiar symbolic elements from the Ministry's communications versus two control groups) and content (logically persuasive versus unpersuasive policy plans). I examine the effects of these manipulations on participants' trust in the policy plan accross the two areas, and test whether residents of the polluted area are less affected by familiar symbolic elements (and more affected by the different in the substantive content). I further supplement this observational comparison between the areas with an experimental treatment for the perceived personal relevance of the policy, and with additional robustness tests.

Contrary to my initial expectations, the results indicate that even high levels of perceived personal relevance do not significantly attenuate the effect of symbols on citizens' trust. These findings tend to suggest that the boundaries of the effect of symbolic communication are wider than expected. I discuss the theoretical and normative implications of these findings.

@@

Symbolic elements, that evoke particular associations and emotions, are included in almost every government communication with citizens, and most most prominently in government advertising and marketing [@eshuis_2012]. These symbols can affect citizens' attitudes unconsciously through a psychological mechanism of evaluative conditioning -- the positive associations towards the symbolic element is transferred to the public organization, following repeated appearance of the symbol in the organization's public communications [@alonbarkat_2017; @dehouwer_2012]. This mechanism is considerred a primitive form of persuasion which requires very little cognitive effort, as opposed to persuasion resulted by thoughtful consideration of substantive arguments (in the administrative context, for instance - evaluating organizations and policies based on performance information). The question under which circumbstances are citizens' more/less susceptible to persuasion via branding, falls within a more general theoretical question of what determines the ways in which individuals process communication messeges.

According to the Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo [@petty_1986; @petty_2011], people may use different processing mechanisms in response to a communication, and accordingly persuasion can occure in different ways.  
depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on a message, they are likely to focus on the arguments in the communication that are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route"). These two routes for persuasion are regarded as two ends of an "elaboration continuum". People may also be located between these two ends of the continuum, and then a mixture of these processes will determine their persuasion.
